### 2017-01-24 manuscript.rmd
reorganized a bit, added to pairwise analyses

Functional genomics, the study of how genomes work, is one of the most rapidly advancing fields in Biology. There is now considerable interest in comparing genome function across species to understand what is conserved, what is variable, and how the diversity of organism traits (including anatomy, physiology, and development) relates to genome diversity. Most published comparisons of genome function across species have, however, neglected to use methods that consider the evolutionary relationships among species. This has two potential negative impacts. First, it can lead to the wrong conclusions about the evolution of genome function. Second, it is a missed opportunity to learn about biology that can only be understood in an explicit historical context that considers these relationships. Here we examine two recently published comparative expression studies that used multiple pairwise comparison rather than phylogenetic approaches to investigate evolutionary patterns of gene expression. We find problems with the pairwise comparisons used in both studies that undermine their published conclusions. This is a concrete demonstration of the inadequacy of pairwise comparisons in comparative genomics studies, and indicates that it will be critical to adopt phylogenetic comparative methods in future analyses.

@@

## Results

@@

The KMRR considers several publicly available datasets. Here we consider their conclusions in the context of Brawand *et al.* 2011 [Brawand:2011du], one of the studies they considered. This analysis of XX different organs across XX mammal species.

For each internal node in each gene tree (Figure 1b), we then measured the phylogenetically independent contrast (PIC) in the expression across this node [@Felsenstein:1985ua]. This is the difference in values of descendent nodes (i.e., the evulutionary change along the branches that descend from the node) scaled by the expected variance (determined largely by the lengths of these branches). These contrasts were then annotated by whether they are made across branches that descend from a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree. Our use of it to look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree is a bit different than the usual application of the method.

@@

@@

@@

While demonstrate a common problem with pairwise comparisons in Levin *et al.* dataset, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study. This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs. This means that there is not a trait that can be mapped onto the tree, as KMRR's tau can be. A full phylogenetic reanalysis would of course be possible be using upstream analysis products to rederive a new expression summary statistic that was suitable for mapping onto a phylogenetic tree.

## Discussion

@@

### Ortholog conjecture

The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). It has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is used to relate findings from well-studied genes to related genes that haven't been investigated in detail. It can be applied to any trait of genes, from gene seqeunce to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it can also be considered a hypothesis about the process of evolution. Specifically, it postulates that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

Despite its importance, there have been relatively few tests of the ortolog conjecture. Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz]. Analyses of domain structure were consistent with ortholog conjecture [@Forslund:2011cw]. There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz].

@@

@@

@@

```{r filtered, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}

  g_age =
  nodes_contrast %>%
    ggplot( aes( x=node_age, y=pic, col=Event ) ) + 
		  geom_point( alpha=0.2 ) + 
		  xlab( "Node age (millions of years)" ) + 
		  ylab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
		  ylim(0,0.3) +
			xlim(0,1000) +
		  ggtitle( "(a)" ) +
		  theme_classic()	
		  #theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )

@@

  var_exp_max = 
    nodes_contrast %>% 
    filter( D=="N" ) %>%
    select( var_exp ) %>%
    max( na.rm=TRUE )

@@

		# Calculate the phenotypic difference between the tips
		D$difference = 
			apply(
				D,
				1,
				function(y) {
					tau_values = nhx@data$Tau[ c( as.integer(y['tip_a']), as.integer( y['tip_b'] ) ) ]
					max( tau_values ) - min( tau_values )
				}
			)

	pairwise_summary %>% 
		filter( ! is.na(D) ) %>%
  	ggplot(aes(x=distance, y=difference, color=D)) + 
  		geom_point(alpha=0.1) +
  		geom_smooth(method=lm, se=FALSE) +
			xlim( 0, 2000 )

>>>>

Functional genomics, the study of how genomes work, is one of the most rapidly advancing fields in Biology. There is now considerable interest in comparing genome function across species to understand what is conserved, what is variable, and how the diversity of organism traits (including anatomy, physiology, and development) relates to genome diversity. Most published comparisons of genome function across species have, however, neglected to use methods that consider the evolutionary relationships among species. This has two potential negative impacts. First, it can lead to the wrong conclusions about the evolution of genome function. Second, it is a missed opportunity to learn about biology that can only be understood in an explicit historical context that considers these relationships. Here we examine two recently published comparative expression studies that used multiple pairwise comparison rather than phylogenetic approaches to investigate evolutionary patterns of gene expression. We find problems with the pairwise comparisons used in both studies that undermine their published conclusions. The first study we examine found that expression tends to be more different between paralogs than orthologs, and interpretted this as evidence of the ortholog conjecture- the hypothesis that there are greater rates of evolutionary change following duplication. Our phylogenetic analysis finds no difference in rates following duplication and speciation. Instead, the greater difference between paralogs is because they tend to be more distantly related to each other than orthologs. The second study claimed to find more variance in gene expression at the midpoint of development than earlier and later when comparing animals from different phyla. We instead find that changes along the branch of a single taxon, the ctenophore, were overcounted becuase they were included in each pairwise comparison. These reanalyses are concrete demonstrations of the inadequacy of pairwise comparisons in comparative genomics studies, and indicates that it will be critical to adopt phylogenetic comparative methods in future analyses.

@@

## Results and Discussion

@@

#### Original analyses

The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). It has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is used to relate findings from well-studied genes to related genes that haven't been investigated in detail. It can be applied to any trait of genes, from gene seqeunce to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it can also be considered a hypothesis about the process of evolution. Specifically, it postulates that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

Despite its importance, there have been relatively few tests of the ortolog conjecture. Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz]. Analyses of domain structure were consistent with ortholog conjecture [@Forslund:2011cw]. There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such study to date.

The KMRR study considered several publicly available datasets. Here we consider their conclusions in the context of Brawand *et al.* 2011 [@Brawand:2011du], one of the studies they considered. This dataset of gene expression for XX different organs across XX mammal species is the best sampled in their reanalyses. Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression. It can range from a value of 0, which indiciates no specifificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates his specificity. It is convenient in that it is a single number of defined range for each gene in each species, though of course since the original expression is multidemensional this means much information is discarded. For example, it does not convey which tissue expression is specific to. Homologous genes could each have a tau value of 1, but one gene could have expression that is specific to the kidney and the other to the brain, for example. Their published analyses are based on pairwise comparisons (Figure 1a) betwen tau for each gene family. They found the correlation for orthologs to be significantly greater than the correlation of tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs. From this concluded that their analyses support the ortholog conjecture. They also went one step further - they concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs."

#### Phylogenetic reanalyses

We reanalyzed the their study using phylogenetic comparative methods, focusing on the dataset of Brawand *et al.* 2011 [Brawand:2011du]. For each internal node in each gene tree (Figure 1b), we measured the phylogenetically independent contrast (PIC) in the expression across this node [@Felsenstein:1985ua]. This is the difference in values of descendent nodes (*i.e.*, the evulutionary change along the branches that descend from the node) scaled by the expected variance (determined largely by the lengths of these branches). These contrasts were then annotated by whether they are made across branches that descend from a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree. Our use of it to look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree is a bit different than the usual application of the method.

@@

```{r ascertainment, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}

  g_age =
  nodes_contrast %>%
    ggplot( aes( x=node_age, y=pic, col=Event ) ) + 
		  geom_point( alpha=0.2 ) + 
		  xlab( "Node age (millions of years)" ) + 
		  ylab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
		  ylim(0,0.3) +
			xlim(0,1000) +
		  ggtitle( "(a)" ) +
		  theme_classic()

	# Examine by expected variance, ie branch length
  var_exp_max = 
    nodes_contrast %>% 
    filter( D=="N" ) %>%
    select( var_exp ) %>%
    max( na.rm=TRUE )

@@

#### Understanding the incongruence between pairwise and phylogenetic methods

		# Add the tip names
		D %<>% mutate ( name_a = nhx@data$G[ D$tip_a ] )
		D %<>% mutate ( name_b = nhx@data$G[ D$tip_b ] )

		# Add the values at the tips
		D %<>% mutate ( tau_a = nhx@data$Tau[ D$tip_a ] )
		D %<>% mutate ( tau_b = nhx@data$Tau[ D$tip_b ] )
		D %<>% mutate ( difference = abs( tau_a - tau_b ) )

		# Calculate the phenotypic difference between the tips with a different method as 
		# check on parsing
		D$difference_old = 
			apply(
				D,
				1,
				function(y) {
					tau_values = nhx@data$Tau[ c( as.integer(y['tip_a']), as.integer( y['tip_b'] ) ) ]
					max( tau_values ) - min( tau_values )
				}
			)

		if ( ! all ( near( D$difference, D$difference_old ) ) ){
			stop( "Inconsistent difference calculations" )
		}

	# Calcualte some summary statistics 
	ortholog_r = 
		pairwise_summary %>% 
		filter( D=="N" ) %>% 
		select( tau_a, tau_b ) %>% 
		cor( ) %>% 
		.[ 1, 2 ]

	paralog_r = 
		pairwise_summary %>% 
		filter( D=="Y" ) %>% 
		select( tau_a, tau_b ) %>% 
		cor( ) %>% 
		.[ 1, 2 ]

	paralog_subset_r = 
		pairwise_summary %>% 
		filter( D=="Y" ) %>% 
		filter( node_age <= max_speciation_age  ) %>% 
		select( tau_a, tau_b ) %>% 
		cor( ) %>% 
		.[ 1, 2 ]

	ortholog_distance_mean = 
		pairwise_summary %>% 
		filter(D=="N") %>% 
		.[["distance"]] %>% 
		mean()

	paralog_distance_mean = 
		pairwise_summary %>% 
		filter(D=="Y") %>% 
		.[["distance"]] %>% 
		mean()

In order to better understand why our phylogenetic analysis seem to support a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their specific result. This is imporant since we are only looking at at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset. We can replicate their result - we find a higher tau correlation between orthologs (R=`r round( ortholog_r, 3) `) then between paralogs (R=`r round( paralog_r, 3)`).

Why is it that pairwise methods and phylogenetic methods lead to the opposite conclusion? The reason is that they describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes. There need not be a different process of evolution following speciation and duplication for there to be different patterns for orthologs and paralogs. That is because there is one other important link between evolutionary process and contemporary patterns - the structure of the tree.

Paralogs could be more different from each other than ortholgs if they tend to be more distantly related to each other than orthologs are. That would provide more time for differences to accumulate between them, regardless of the history of duplication and speciation.  This is, in fact, what we find in this dataset. While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` years. This is at least in part because, while the maximum distance between orthogs is two times the depth of the oldest speciation event in the tree, many duplications precede this oldest sampled speciation event. The distance between paralogs resulting from these deeper duplications can be much larger.

If we consider only the paralogs that arise from the duplication events the same age or younger than the oldest speciation event, the observed pattern is very different. The correlation of tau between these shallower paralog pairs, R=`r round( paralog_subset_r, 3)`, is much stronger than for the correlation between all paralog pairs, R=`r round( paralog_r, 3)`. This is much closer to the ortholog comparison, which has R=`r round( ortholog_r, 3) `.

The other issue with the pairwise comparisons is that many are summarizing changes along many branches in the phylogeny, not just the branches that follow the divergence event. Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and cnages along all these branches will impact the final pattern. The phylogenetic methods isolate the changes under consideration. There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss. But these affect both methods. The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches.

The reason that paralogs pairs have tau values that are more different than orthog pairs is that they tend to be more distantly related, not because there are differences in evolutionary rates of change in tau following duplication and speciation events. This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distintly related the two genes are. This has several implications. One is that it is an example of the limitations of the terms orthology and paralogy [@Dunn:2016er].

@@

While we do demonstrate a common problem with pairwise comparisons in the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study. This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs. This means that there is not a trait that can be mapped onto the tree, as KMRR's tau can be. A full phylogenetic reanalysis would of course be possible be using upstream analysis products to rederive a new expression summary statistic that was suitable for mapping onto a phylogenetic tree.

@@

@@

Studies of evolutionary functional genomics should not be focused on the tips of the tree, they should explicitly delve into the tree itself.

@@

We considered only the data from Brawand *et al.* 2011 [@Brawand:2011du]. KMRR examined multiple subsets of the data. We left in sex chromosome genes and testes expression data, as this corresponded to the analyses that provided the strongest support for the ortholog conjecture and therefore the most conservative reconsideration of it.

@@

```{r ascertainment_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}

@@

@@

```{r pairwise_plot, echo=FALSE, cache=TRUE, message=F, warning=F}	
	pairwise_summary %>% 
		filter( ! is.na(D) ) %>%
  	ggplot(aes(x=distance, y=difference, color=D)) + 
  		geom_point(alpha=0.1) +
  		geom_smooth(method=lm, se=FALSE) +
			xlim( 0, 2000 )
